Ravi Sood
March 13, 2014
BIOST 578A
Homework 3
========================================================

1. Get the data from GEO. Note: used code from class notes.
```{r}
# Load necessary packages
library(GEOquery)
library(limma)
library(data.table)
library(pheatmap)
library(GSEABase)

# Download the supplementary files
getGEOSuppFiles("GSE45735", makeDirectory=FALSE, baseDir = "Data/GEO/")

# Note: prior to using read.table, manually unzipped the text files and removed extraneous rows from T14 txt file
files <- list.files(path = "Data/GEO/", pattern = "GSE45735_T.*.txt", full.names = TRUE)
file_list <- lapply(files, read.table, header=TRUE)

# Remove duplicated rows
file_list_unique <- lapply(file_list, function(x){x<-x[!duplicated(x$Gene),]; 
                                                  x <- x[order(x$Gene),]; 
                                                  rownames(x) <- x$Gene;
                                                  x[,-1]})

# Take the intersection of all genes
gene_list <- Reduce(intersect, lapply(file_list_unique, rownames))
file_list_unique <- lapply(file_list_unique, "[", gene_list,)
matrix <- as.matrix(do.call(cbind, file_list_unique))

# Obtain the pData from GEO
gd <- getGEO("GSE45735", destdir = "Data/GEO/")
pd <- pData(gd[[1]])

# Clean up the pData
pd_small <- pd[!grepl("T13_Day8",pd$title),] # Remove row corresponding to T13 Day8
pd_small$day <- sapply(strsplit(gsub(" \\[PBMC\\]", "", pd_small$title),"_"),"[",2) # Create column for "day"
pd_small$subject <- sapply(strsplit(gsub(" \\[PBMC\\]", "", pd_small$title),"_"),"[",1) # Create column for "subject"
colnames(matrix) <- rownames(pd_small)
```

2. Use voom and limma to find genes that are differentially expressed at each time point compared to baseline (day 0). Use an FDR cutoff of 0.01. Display your results using pheatmap showing the log fold-change of the differentially expressed genes grouped by time point.
```{r}
# Create ExpressionSet
eset_hw3 <- ExpressionSet(assayData = matrix+1)
pData(eset_hw3) <- pd_small

# Create design matrix
design_hw3 <- model.matrix(~subject+day, eset_hw3)
voom_hw3 <- voom(eset_hw3, design = design_hw3)

# Fit linear model
lm <- lmFit(voom_hw3, design_hw3)

# Calculate empirical Bayes statistics 
eb <- eBayes(lm)

# Find DE genes (FDR = 0.01) at each timepoint and store as list of data tables whose list index corresponds to time point being compared to baseline
FDR <- 0.01
days <- 1:10
tt_list <- vector("list", length(days))
for (i in days) {
  coef <- paste0("dayDay",i)
  df <- topTable(eb, coef=coef, number=Inf, sort.by="none")
  tt_list[[i]] <- as.data.table(df)
  tt_list[[i]]$significant <- ifelse(tt_list[[i]]$adj.P.Val<FDR, "Yes", "No")
  tt_list[[i]]$gene <- rownames(df)
  setkey(tt_list[[i]], gene)
}

# Create variable indicating whether gene is differentially expressed at any time point
tt_list[[1]]$anytime <- ifelse(tt_list[[1]]$significant=="Yes", "Yes", "No")
for (i in days[-1]) {
    tt_list[[1]]$anytime <- ifelse(tt_list[[1]]$significant=="No" & tt_list[[i]]$significant=="Yes", "Yes", tt_list[[1]]$anytime)
}

# Create data frame of logFC values relative to day 0 at each time point
setnames(tt_list[[1]],"logFC","Day 1")
hm_dt <- tt_list[[1]][, c("AveExpr","t","P.Value","adj.P.Val","B","significant"):=NULL]
for (i in 2:length(days)) {
  setnames(tt_list[[i]], "logFC", paste("Day",i))
  tt_list[[i]][, c("AveExpr","t","P.Value","adj.P.Val","B","significant"):=NULL]
  hm_dt <- merge(hm_dt, tt_list[[i]], all.x=TRUE)
}
hm_df <- data.frame(hm_dt)
rownames(hm_df) <- hm_dt$gene
colnames(hm_df) <- colnames(hm_dt)

# Delete rows corresponding to genes not differentially expressed at any time point
hm_df <- hm_df[hm_df$anytime=="Yes",]

# Delete extraneous columns and plot heatmap
hm_df$gene <- NULL
hm_df$anytime <- NULL
hm_m <- data.matrix(hm_df)
pheatmap(hm_m,cluster_cols=FALSE,scale="row") 
```

3. Perform a GSEA analysis using camera and the MSigDB Reactome pathway gene signatures. Display your results using pheatmap, again group by timepoint. This is similar to what we've done in class.
```{r}
# Obtain gene indices for camera [Note: first had to manually download MSigDB Reactome gene sets]
c2_set_hw3 <- getGmt("GSEA Sets/c2.cp.reactome.v4.0.symbols.gmt")
gene_ids_hw3 <- geneIds(c2_set_hw3)
sets_indices_hw3 <- symbols2indices(gene_ids_hw3, rownames(eset_hw3))

# Find enriched gene sets over time
desets_list <- vector("list", length(days))
subjects <- length(unique(pData(eset_hw3)$subject))
for (i in days) {
  cont <- paste0("dayDay",i)
  cont_matrix <- makeContrasts(cont, levels=design_hw3)
  desets_list[[i]] <- camera(voom_hw3, index=sets_indices_hw3, design=design_hw3, contrast=cont_matrix)
}

# Draw heatmap of enriched gene sets over time
PValue <- sapply(desets_list, function(x){ifelse(x$Direction=="Up", -10*log10(x$PValue), 10*log10(x$PValue))})
rownames(PValue) <- rownames(desets_list[[1]])
PValue_max <- rowMax(abs(PValue))
PValue_small <- PValue[PValue_max>30, ]
anno <- data.frame(Time=paste0("Day",days))
rownames(anno) <- colnames(PValue_small)  <- paste0("Day",days)
pheatmap(PValue_small, cluster_cols=FALSE, scale="row")
```